---
title: "Introduction to the Stan Language"
author: "Ben Goodrich"
date: "`r format(Sys.time(), '%B %d, %Y')`"
autosize: true
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amssymb}
   - \usepackage{color}
output:
  ioslides_presentation:
    widescreen: yes
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
slides > slide:not(.nobackground):after {
  content: '';
}
.forceBreak { -webkit-column-break-after: always; break-after: column; }
</style>

```{r setup, include=FALSE}
options(width = 90)
library(knitr)
knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar = c(4, 4, .1, .1), las = 1)  # smaller margin on top and right
})
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
  lines <- options$output.lines
  if (is.null(lines)) {
    return(hook_output(x, options))  # pass to default hook
  }
  x <- unlist(strsplit(x, "\n"))
  more <- "..."
  if (length(lines)==1) {        # first n lines
    if (length(x) > lines) {
      # truncate the output, but add ....
      x <- c(head(x, lines), more)
    }
  } else {
    x <- c(more, x[lines], more)
  }
  # paste these lines together
  x <- paste(c(x, ""), collapse = "\n")
  hook_output(x, options)
})
library(rstan)
options(mc.cores = 4L)
rstan_options(auto_write = TRUE)
```

## 2014 Ebola Crisis

* In 2014 there was an(other) outbreak of Ebola in Africa
* $n = 7$ western medical professionals were infected and given an experimental drug called ZMapp.
  Goal is to decide whether ZMapp is effective using Stan.
* The binomial distribution is standard for evaluating the probability of $y$ successes in $n$
  independent trials with common success probability $\pi$
  $$\Pr\left(y \mid \pi, n\right) = {n \choose y} \pi^y \left(1 - \pi\right)^{n - y}$$
* For OBSERVED $y$ (and $n$), we can write a log-likelihood function of $\pi$ as
  $$\ell\left(\pi\right) = y \ln \pi + \left(n - y\right) \ln \left(1 - \pi\right)$$

## Posterior Distribution in Explicit Form

* A Beta distribution is often used as a prior for probabilities, which has the PDF
  $$f\left(\pi \mid a, b\right) = B\left(a,b\right)^{-1} \pi^{a - 1} \left(1 - \pi\right)^{b - 1}$$
  where $B\left(a,b\right) = \int_0^1 x^{a - 1} \left(1 - x\right)^{b - 1} dx$ is the Beta function
* Bayes Rule for a continuous $\pi$ can be written as a PDF
  $$\begin{eqnarray}
  f\left(\pi \mid y, n, a, b\right) & = & \frac{f\left(\pi \mid a, b\right) \mathrm{Pr}\left(y \mid \pi, n\right)}
  {\mathrm{Pr}\left(y \mid n, a, b\right)} \\
  & = & \frac{B\left(a,b\right)^{-1}
  \pi^{a - 1} \left(1 - \pi\right)^{b - 1} {n \choose y} \pi^y \left(1 - \pi\right)^{n - y}}
  {\mathrm{Pr}\left(y \mid n, a, b\right)} \\
  & \propto & \pi^{a^\ast - 1} \left(1 - \pi\right)^{b^\ast - 1}
  \end{eqnarray}$$
  where $a^\ast = a + y$ and $b^\ast = b + n - y$
  
## Workflow for Stan

- You write the program in a (text) .stan file with a C-like syntax
- Stan's parser, `stanc`, does three things:
    - checks that program is syntactically valid and tells you if not
    - writes a conceptually equivalent C++ source file to disk
    - C++ compiler creates a binary file from the C++ source
- When you have some C++ like `x = mu + sigma * z;`
    - C++ can automatically store $\frac{\partial x}{\partial\mu}$, $\frac{\partial x}{\partial\sigma}$,
    and $\frac{\partial x}{\partial z}$ by overloading arithmetic operators and handle the chain-rule for you
    - Called automatic differentiation (not numerical differentiation)
    - Unless $\mu$, $\sigma$, or $z$ is constant, in which case it does not bother
- You execute the binary from an interface
- You analyze the resulting samples from the posterior

## Primitive Object Types in Stan

- In Stan / C++, variables must declared with types
- In Stan / C++, statements are terminated with semi-colons
- Primitive scalar types: `real x;` or `int K;`
    - Unknowns cannot be `int` because no derivatives and hence no HMC
    - Can condition on integer data because no derivatives are needed
- Implicitly real `vector[K] z;` or `row_vector[K] z;`
- Implicitly real `matrix[N,K] X;` can have 1 column / row
- Arrays are just holders of any other homogenous objects
    - `real x[N]` is similar to `vector[N] x;` but lacks linear algebra functions
    - `vector[N] X[K];` and `row_vector[K] X[N]` are similar to
      `matrix[N,K] X;` but lack linear algebra functionality, although
      they have uses in loops
- Vectors and matrices cannot store integers,
  so instead use possibly multidimensional integer arrays `int y[N];` or `int Y[N,P];`

## The `lookup` Function in **rstan**

- Can input the name of an R function, in which case it will try
to find an analagous Stan function
- Can input a regular expression, in which case it will find matching
Stan functions that match
```{r, size='footnotesize',comment="#", message = FALSE}
rstan::lookup("^inv.*[^gf]$") # functions starting with inv but not ending with g or f
```

## Optional `functions` Block of Stan Programs

- Stan permits users to define and use their own functions
- If used, must be defined in a leading `functions` block
- Can only validate constraints inside user-defined functions
- Very useful for several reasons:
    - Easier to reuse across different .stan programs
    - Makes subsequent chunks of code more readable
    - Enables posteriors with Ordinary Differential Equations, algebraic
      equations, and integrals
    - Can be exported to R via `expose_stan_functions()`
- All functions, whether user-defined or build-in, must be called by
argument position rather than by argument name, and there are no default
arguments
- User-defined functions cannot have the same name as existing functions
or keywords and are case-sensitive

## Illustration of `functions` Block

```{r, echo = FALSE, comment = ""}
writeLines(readLines("hello.stan")[1:19])
```

## Constrained Object Declarations in Stan

Outside of the `functions` block, any primitive object can have bounds:

- `int<lower = 1> K;` `real<lower = -1, upper = 1> rho;`
- `vector<lower = 0>[K] alpha;` and similarly for a `matrix`
- A `vector` (but not a `row_vector`) can be further specialized:
    - `unit_vector[K] x;` implies $\sum_{k=1}^{K}x_{k}^{2}=1$
    - `simplex[K] x;` implies $x_{k}\geq0\,\forall k$ and $\sum_{k=1}^{K}x_{k}=1$
    - `ordered[K] x;` implies $x_{j}<x_{k}\,\forall j<k$
    - `positive_ordered[K] x;` implies $0<x_{j}<x_{k}\,\forall j<k$
- A `matrix` can be specialized to enforce constraints:
    - `cov_matrix[K] Sigma;` or better `cholesky_factor_cov[K, K] L;`
    - `corr_matrix[K] Lambda;` or `cholesky_factor_corr[K] C;`

## `data` and `transformed data` Blocks{.columns-2}
### "Required" `data` Block

- Contains all knowns passed to Stan as a NAMED list in R or as a dict in Python, such as
  outcomes $\left(\mathbf{y}\right)$, covariates $\left(\mathbf{X}\right)$,
  constants $\left(K\right)$, and / or hyperparameters $\left(a\right)$
- Basically, everything posterior distribution conditions on
- Can have comments in C++ style (`//` or `/* ... */`)
- Whitespace is essentially irrelevant, except after keywords
- Symbols are evaluated sequentially

<p class="forceBreak"></p>

### Optional `transformed data` Block

- Is executed only once before the iterations start
- Used to calculate needed functions of data
- Not necessary if calling Stan from R or Python with everything in `data`
- Can use it to check that data was passed correctly from R
- Need to declare objects before they can be assigned (=) but can be on the same line
- All declarations must come directly after the opening `{`

## Illustration of `data` and `transformed data` Blocks

```{r, echo = FALSE, comment = ""}
writeLines(readLines("hello.stan")[21:38])
```

## `parameters` and `transformed parametes` Blocks{.columns-2}
### "Required" `parameters` Block

- Declare exogenous unknowns for desired posterior distribution
- Cannot declare any integer parameters, only reals and containers thereof
- Must specify the parameter space but lower and upper
bounds are implicitly $\pm\infty$ if unspecified
- The change-of-variables adjustment due to the transformation from
an unconstrained parameter space to the (in this case, positive) constrained space
is handled automatically and added to `target`

<p class="forceBreak"></p>

### Optional `transformed parameters`

- Similar in structure to the `transformed data` block but is executed every leapfrog step
- All declarations must come directly after the opening `{`
- Calculate endogenous unknowns that are deterministic functions of things declared in earlier blocks
- Used to create interesting intermediate inputs to the log-kernel
- Declared constraints are validated and samples are stored
- Often used in multilevel models to define group-specific unknowns

## "Required" `model` Block

- Can declare endogenous unknowns, assign to them, and use them
- All declarations must come directly after the opening `{`
- Constraints cannot be declared / validated and samples not stored
- The `model` block must define (something proportional to) $\text{target}=\log\left(f\left(\boldsymbol{\theta}\right)\times f\left(\left.\mathbf{y}\right|\boldsymbol{\theta},\cdot\right)\right)=\log f\left(\boldsymbol{\theta}\right)+\log f\left(\left.\mathbf{y}\right|\boldsymbol{\theta},\cdot\right)$
- There is an internal reserved symbol called `target` that is
initialized to zero (before change-of-variable adjustments) you increment by `target += ...;`
- Functions ending `_lpdf` or `_lpmf` return scalars even if some of their arguments are vectors or 
  one-dimensional arrays, in which case it sums the log density/mass over the presumed conditionally
  independent elements

## Optional `generated quantities` Block

- Can declare more endogenous knowns, assign to them, and use them
- Samples are stored in the output
- Can reference anything except stuff in the `model` block
- Can also do this in R or Python afterward, but primarily used for
    - Interesting functions of posterior that do not involve likelihood
    - Posterior predictive distributions and / or functions thereof
    - The log-likelihood for each observation to pass to `loo`

## Illustration of Remaining Blocks

```{r, echo = FALSE, comment = ""}
writeLines(readLines("hello.stan")[-(1:38)])
```

## Calling `stan` in `library(rstan)`

```{r, hello, cache = TRUE, results = "hide", message = FALSE}
hello <- stan("hello.stan", data = list(n = 7, y = 5, a = 4, b = 2), cores = 4)
```
```{r}
dim(hello) # 1000 draws on each of 4 chains for 4 quantities
print(hello, digits = 2) # lp__ is just target from the Stan program
```

