---
title: "Stan Language Lab Answers"
author: "Ben Goodrich"
editor_options:
  chunk_output_type: console
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amssymb}
   - \usepackage{color}
output: 
  html_document: 
    number_sections: yes
---

The questions and additional context are also [available](afternoon_lab.Rmd).

```{r, message = FALSE}
US_Open2010 <- readr::read_csv("US_Open2010.csv")
```

# Simplest Model

## Simplest `model` Block

```{r, comment = "", echo = FALSE}
writeLines(readLines("simplest.stan"))
```

## Running the Simplest Stan Program

Among professional bowlers, a strike is the most common outcome and they almost never roll a gutter
ball or only knock down a few pins on their first roll. Thus, the $\mathbf{a}$ vector should be
of size $11$, positive, and generally increasing. Perhaps something like

```{r}
a <- c(2, 1, 1, 1, 1, 2, 4, 8, 16, 32, 64)
```
which implies a prior expectation for $\boldsymbol{\pi}$ of

```{r}
round(a / sum(a), digits = 3)
```

Then, we can update our prior beliefs about $\boldsymbol{\pi}$ by conditioning on the data
```{r, message = FALSE}
library(rstan)
options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

post <- stan("simplest.stan", data = list(N = nrow(US_Open2010), roll_1 = US_Open2010$roll_1 + 1,
                                          roll_2 = US_Open2010$roll_2 + 1, a = a), refresh = 0)
print(post, pars = "pi", digits = 3)
```

## Simplest `generated quantities` Block

The `generated quantities` block is included in simplest.stan .

# Intermediate Model

## Intermediate `model` Block

```{r, comment = "", echo = FALSE}
writeLines(readLines("intermediate.stan"))
```

## Intermediate `functions` and `generated quantities` Block

The game_rng.stan file can be written as

```{r, comment = "", echo = FALSE}
writeLines(readLines("game_rng.stan"))
```

## Running the Intermediate Stan Program

```{r}
post <- stan("intermediate.stan", data = list(N = nrow(US_Open2010), roll_1 = US_Open2010$roll_1 + 1,
                                              roll_2 = US_Open2010$roll_2 + 1, a = a), refresh = 0)
post
```
The posterior distribution of a professional bowler's score has a mean and median near $200$
with a standard deviation of about $26$. This is unreasonably low for a professional bowler, so
what do you think is wrong about this model?

# Multilevel Stan Program

## Multilevel Stan Program

```{r, comment = "", echo = FALSE}
writeLines(readLines("multilevel.stan"))
```

## Running the Multilevel Stan Program

```{r, multilevel, cache = TRUE, warning = FALSE, message = FALSE}
# drop subsequent rounds
US_Open2010 <- US_Open2010[!duplicated(paste0(US_Open2010$Bowler, "_", US_Open2010$Frame)), ]
J <- length(unique(US_Open2010$Bowler)) # number of bowlers
roll_1 <- t(matrix(US_Open2010$roll_1 + 1, nrow = 10, ncol = J))
roll_2 <- t(matrix(US_Open2010$roll_2 + 1, nrow = 10, ncol = J))

post <- stan("multilevel.stan",
             data = list(J = J, roll_1 = roll_1, roll_2 = roll_2, a = a, rate = 1), 
             refresh = 0, control = list(adapt_delta = 0.9))
```
Setting `adapt_delta` to $0.9$ (its default value is $0.8$) reduces the
warnings about divergent transitions, which are important. For more information, see this 
[FAQ](https://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup).

```{r}
winner <- rstan::extract(post, pars = "winner")[[1]]
```
is a vector of size $4000$ of the bowler with the highest score. Based on
```{r}
prop.table(table(winner))
```
bowler number 16 has the highest probability of winning the bowling tournament,
but only about $0.15$
