---
title: "Bayesian Probability Lab Answers"
author: "Ben Goodrich"
editor_options:
  chunk_output_type: console
output: 
  html_document: 
    number_sections: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(crayon)
card <- function(value = c("A", as.character(2:9), "10", "J", "Q", "K"), 
                 suit = c("S", "H", "D", "C"),
                 style = NULL) {
  value <- match.arg(value)
  suit <- match.arg(suit)
  
  deck <- list(S = c("\U1F0A1", "\U1F0A2", "\U1F0A3", "\U1F0A4", "\U1F0A5", "\U1F0A6", "\U1F0A7",
                     "\U1F0A8", "\U1F0A9", "\U1F0AA", "\U1F0AB", "\U1F0AD", "\U1F0AE"),
               H = c("\U1F0B1", "\U1F0B2", "\U1F0B3", "\U1F0B4", "\U1F0B5", "\U1F0B6", "\U1F0B7",
                     "\U1F0B8", "\U1F0B9", "\U1F0BA", "\U1F0BB", "\U1F0BD", "\U1F0BE"),
               D = c("\U1F0C1", "\U1F0C2", "\U1F0C3", "\U1F0C4", "\U1F0C5", "\U1F0C6", "\U1F0C7",
                     "\U1F0C8", "\U1F0C9", "\U1F0CA", "\U1F0CB", "\U1F0CD", "\U1F0CE"),
               C = c("\U1F0D1", "\U1F0D2", "\U1F0D3", "\U1F0D4", "\U1F0D5", "\U1F0D6", "\U1F0D7",
                     "\U1F0D8", "\U1F0D9", "\U1F0DA", "\U1F0DB", "\U1F0DD", "\U1F0DE"))
               
  value <- switch(value,
                  "A" = 1L,
                  "10" = 10L,
                  "J" = 11L,
                  "Q" = 12L,
                  "K" = 13L,
                  as.integer(value))
  if (is.null(style)) style <- if (suit %in% c("H", "D")) red else reset
  return(fansi::sgr_to_html(style(deck[[suit]][value])))
}
```

The questions and additional context are also [available](morning_lab.Rmd).

# Byron Kaverman's Decision

## Prior

The _ex ante_ probability of being dealt `r card("A", "C")` `r card("5", "C")` can be calculated as

$$\frac{1}{52} \times \frac{1}{51} \times 2 = \frac{1}{1326}$$

because there is a $1$ in $52$ chance of being dealt `r card("A", "C")`, which leaves $51$ cards
left in the deck including the only `r card("5", "C")`. But the order that these two cards are
dealt can also be reversed, which accounts for the factor of $2$. Alternatively, any two-card
hand has a 

$$\frac{1}{52 \choose 2} = \frac{2!\left(52 - 2\right)!}{52!} = \frac{2}{52\times 51} = \frac{1}{1326}$$

probability of being dealt.


## Posterior

If $A$ represents the event that Kaverman is dealt dealt `r card("A", "C")` `r card("5", "C")`, we know
from the previous problem that its prior probability is

$$\Pr\left(A\right) = \frac{1}{1326}$$

Then, if $B$ represents the event that Kaverman goes all-in, we can write the probability of that Kaverman 
goes all-in with `r card("A", "C")` `r card("5", "C")` as

$$\Pr\left(B \mid A\right) = 1$$

Finally, we need the denominator of Bayes' Rule, which is the _ex ante_ probability of going all-in, which can
be written as

$$\Pr\left(B\right) = \frac{138}{1326}$$

because there are $138$ hands enumerated where the expectation of going all-in is better than the expectation
of folding for someone in Kaverman's situation. Thus, Bayes' Rule simplifies to

$$\Pr\left(A \mid B\right) = \frac{\Pr\left(A \right) \Pr\left(B \mid A \right)}{\Pr\left(B \right)} = 
\frac{1 / 1326 \times 1}{138 / 1326} = \frac{1}{138} \approx 0.007$$

Due to the fact that we are essentially dealing with discrete uniform distributions, you could have arrived at 
$\Pr\left(A \mid B\right) = \frac{1}{138}$ by inspection, as opposed to explicitly evaluating the right-hand side 
of Bayes' Rule. But it is nice to know that this shortcut that is applicable in the discrete uniform case is 
actually a special case of Bayes' Rule, which as we saw in the bowling example does not rely on uniformity.

# Fedor Holz's First Decision

## Probability of Three of a Kind

To answer this question, we can use the function from the previous lecture to evaluate the PMF of the 
hypergeometric distribution:
```{r}
dhyper <- function(x, N, K, n) choose(K, x) * choose(N - K, n - x) / choose(N, n)
```

We are looking for the probability that among the five face-up community cards in the middle, there is one 10
(so that Holz makes three of a kind) and, _given that there is a 10 among those five cards_, the probability 
that Kaverman does not make a higher three of a kind. After pulling out Kaverman's two cards and Holz's two cards, 
there are $48$ cards left in the deck, so this probability is given by

```{r}
dhyper(x = 1, N = 48, K = 2, n = 5) *   # probability Holz makes three of a kind
  (1 - dhyper(1, N = 47, K = 2, n = 4)) # conditional probability Kaverman does not make three of a kind
```

So, in this situation Kaverman's hand would be a strong favorite to beat Holz's hand, and this conclusion
would still hold if we were more rigorous about accounting for the tiny probability that one or both of
their hands ends up being better than three of a kind.

## Probability of Trailing

This question can be answered just by counting up the proportion of hands that Kaverman would go all-in with
that are a favorite against Holz's `r card("10", "S")` `r card("10", "C")`. To do that, we eliminate hands
involving the `r card("10", "S")` or `r card("10", "C")` from the $138$ hands that Kaverman would go all-in 
with, leaving 

* Big pairs (Js or better), of which there are four pairs $\times$ six suit combinations for 24 hands:
    * `r card("A", "S")` `r card("A", "H")`, `r card("A", "S")` `r card("A", "D")`, `r card("A", "S")` `r card("A", "C")`,
      `r card("A", "H")` `r card("A", "D")`, `r card("A", "H")` `r card("A", "C")`, `r card("A", "D")` `r card("A", "C")`
    * `r card("K", "S")` `r card("K", "H")`, `r card("K", "S")` `r card("K", "D")`, `r card("K", "S")` `r card("K", "C")`,
      `r card("K", "H")` `r card("K", "D")`, `r card("K", "H")` `r card("K", "C")`, `r card("K", "D")` `r card("K", "C")`  
    * `r card("Q", "S")` `r card("Q", "H")`, `r card("Q", "S")` `r card("Q", "D")`, `r card("Q", "S")` `r card("Q", "C")`,
      `r card("Q", "H")` `r card("Q", "D")`, `r card("Q", "H")` `r card("Q", "C")`, `r card("Q", "D")` `r card("Q", "C")`
    * `r card("J", "S")` `r card("J", "H")`, `r card("J", "S")` `r card("J", "D")`, `r card("J", "S")` `r card("J", "C")`,
      `r card("J", "H")` `r card("J", "D")`, `r card("J", "H")` `r card("J", "C")`, `r card("J", "D")` `r card("J", "C")`
* A pair of 10s, of which there is only $1$ combination left:
    * `r card("10", "H")` `r card("10", "D")`
* Medium pairs (6s or better), of which there are four pairs $\times$ six suit combinations for 24 hands:
    * `r card("9", "S")` `r card("9", "H")`, `r card("9", "S")` `r card("9", "D")`, `r card("9", "S")` `r card("9", "C")`,
      `r card("9", "H")` `r card("9", "D")`, `r card("9", "H")` `r card("9", "C")`, `r card("9", "D")` `r card("9", "C")`,  
    * `r card("8", "S")` `r card("8", "H")`, `r card("8", "S")` `r card("8", "D")`, `r card("8", "S")` `r card("8", "C")`,
      `r card("8", "H")` `r card("8", "D")`, `r card("8", "H")` `r card("8", "C")`, `r card("8", "D")` `r card("8", "C")`,
    * `r card("7", "S")` `r card("7", "H")`, `r card("7", "S")` `r card("7", "D")`, `r card("7", "S")` `r card("7", "C")`,
      `r card("7", "H")` `r card("7", "D")`, `r card("7", "H")` `r card("7", "C")`, `r card("7", "D")` `r card("7", "C")`,  
    * `r card("6", "S")` `r card("6", "H")`, `r card("6", "S")` `r card("6", "D")`, `r card("6", "S")` `r card("6", "C")`,
      `r card("6", "H")` `r card("6", "D")`, `r card("6", "H")` `r card("6", "C")`, `r card("6", "D")` `r card("6", "C")`
* Ace and at least a nine of the same suit, of which there are five intersections for each of the red suits 
  and four remaining intersections for each of the black suits for a total of 18 hands:
    * `r card("A", "S")` `r card("K", "S")`, `r card("A", "S")` `r card("Q", "S")`, `r card("A", "S")` `r card("J", "S")`,
      `r card("A", "S")` `r card("9", "S")`
    * `r card("A", "H")` `r card("K", "H")`, `r card("A", "H")` `r card("Q", "H")`, `r card("A", "H")` `r card("J", "H")`,
      `r card("A", "H")` `r card("10", "H")`, `r card("A", "H")` `r card("9", "H")`
    * `r card("A", "D")` `r card("K", "D")`, `r card("A", "D")` `r card("Q", "D")`,
      `r card("A", "D")` `r card("J", "D")`, `r card("A", "D")` `r card("10", "D")`, `r card("A", "D")` `r card("9", "D")`
    * `r card("A", "C")` `r card("K", "C")`, `r card("A", "C")` `r card("Q", "C")`, `r card("A", "C")` `r card("J", "C")`,
      `r card("A", "C")` `r card("9", "C")`
* Ace and a five of the same suit, for any of the four suits: `r card("A", "S")` `r card("5", "S")`,
  `r card("A", "H")` `r card("5", "H")`, `r card("A", "D")` `r card("5", "D")`, `r card("A", "C")` `r card("5", "C")`
* Both cards at least 10 of the same suit, of which there are six intersections for each of the red suits
  and three remaining intersections for each of the black suits, leaving 18 hands:
    * `r card("K", "S")` `r card("Q", "S")`, `r card("K", "S")` `r card("J", "S")`,
      `r card("Q", "S")` `r card("J", "S")`
    * `r card("K", "H")` `r card("Q", "H")`, `r card("K", "H")` `r card("J", "H")`, `r card("K", "H")` `r card("10", "H")`,
      `r card("Q", "H")` `r card("J", "H")`, `r card("Q", "H")` `r card("10", "H")`, `r card("J", "H")` `r card("10", "H")`  
    * `r card("K", "D")` `r card("Q", "D")`, `r card("K", "D")` `r card("J", "D")`, `r card("K", "D")` `r card("10", "D")`,
      `r card("Q", "D")` `r card("J", "D")`, `r card("Q", "D")` `r card("10", "D")`, `r card("J", "D")` `r card("10", "D")` 
    * `r card("K", "C")` `r card("Q", "C")`, `r card("K", "C")` `r card("J", "C")`,
      `r card("Q", "C")` `r card("J", "C")`
* Ace and at least a Jack of different suit, of which there are six intersections $\times {4 \choose 2}$ suits for 36 hands:
    * `r card("A", "S")` `r card("K", "H")`, `r card("A", "S")` `r card("K", "D")`, `r card("A", "S")` `r card("K", "C")`
    * `r card("A", "S")` `r card("Q", "H")`, `r card("A", "S")` `r card("Q", "D")`, `r card("A", "S")` `r card("Q", "C")`
    * `r card("A", "S")` `r card("J", "H")`, `r card("A", "S")` `r card("J", "D")`, `r card("A", "S")` `r card("J", "C")`
    * `r card("A", "H")` `r card("K", "S")`, `r card("A", "H")` `r card("K", "D")`, `r card("A", "H")` `r card("K", "C")`  
    * `r card("A", "H")` `r card("Q", "S")`, `r card("A", "H")` `r card("Q", "D")`, `r card("A", "H")` `r card("Q", "C")`
    * `r card("A", "H")` `r card("J", "S")`, `r card("A", "H")` `r card("J", "D")`, `r card("A", "H")` `r card("J", "C")`
    * `r card("A", "D")` `r card("K", "S")`, `r card("A", "D")` `r card("K", "H")`, `r card("A", "D")` `r card("K", "C")`
    * `r card("A", "D")` `r card("Q", "S")`, `r card("A", "D")` `r card("Q", "H")`, `r card("A", "D")` `r card("Q", "C")`  
    * `r card("A", "D")` `r card("J", "S")`, `r card("A", "D")` `r card("J", "H")`, `r card("A", "D")` `r card("J", "C")`
    * `r card("A", "C")` `r card("K", "S")`, `r card("A", "C")` `r card("K", "H")`, `r card("A", "C")` `r card("K", "D")`
    * `r card("A", "C")` `r card("Q", "S")`, `r card("A", "C")` `r card("Q", "H")`, `r card("A", "C")` `r card("Q", "D")`
    * `r card("A", "C")` `r card("J", "S")`, `r card("A", "C")` `r card("J", "H")`, `r card("A", "C")` `r card("J", "D")`

which amounts to $125$ hands that Kaverman could have, of which Kaverman would be a favorite only if he had one of the
$24$ hands with a big pair. Also, neither is a favorite if Kaverman has `r card("10", "H")` `r card("10", "D")`. So, 
the probability that Holz's hand is not a favorite to beat Kaverman's hand is

$$\frac{25}{125} = \frac{1}{5}$$

Excluding some minor strategy considerations, this is essentially why the commentator says at 0:54 of the video that 
Holz "is obviously calling".

# Rick Salomon's Decision

We saw above that Kaverman would go all-in with approximately $10.4\%$ hands he could have been dealt. By calling,
Holz is essentially saying his hand is better than the average of those $10.4\%$ hands, which is to say that
Holz's hand is roughly among the best $5.2\%$ of hands he could have been dealt. By subsquently going all-in
Salomon is essentially saying that his hand is better than the average of the hands that Kaverman and
Holz could both have, given that Kaverman went all-in and Holz called. Since there are three players involved,
an exact calculation could get unnecessarily complicated, but Salomon is primarily worried about Holz calling
his $18,800,000$ chip raise over Kaverman's original $8,025,000$ chip bet. Salomon would be comfortable with that 
if his hand were better than the average of the $5.2\%$ of hands Holz would call Kaverman's all-in bet with. Thus, Salomon
is essentially saying that his hand is among the best $2.6\%$ hands he could have been dealt, which is
$34$ or $35$ hands out of ${52 \choose 2} = 1326$. Not coincidentally, that is about equal to the $24$ hands
with a pair of Aces, Kings, Queens, or Jacks, plus the $12$ combinations of an Ace and a King that Salomon
could have been dealt.

# Fedor Holz's Second Decision

## Beating a King that is not `r card("K", "H")`

To answer this question, we need to again evaluate the PMF of the hypergeometric distribution. It is easier
to determine the probability that Salomon's hand beats Holz's hand and to subtract that from one. The (most)
relevant considerations are 

* The probability that Salomon improves to one pair when the five community cards include exactly one Ace or one King 
```{r}
Pr_1pair <- dhyper(x = 1, N = 48, K = 6, n = 5)
```
* The probability that Salomon improves to two pair when the five community cards include exactly one Ace and one King
```{r}
Pr_2pair <- dhyper(x = 1, N = 48, K = 3, n = 5) * dhyper(x = 1, N = 47, K = 3, n = 4)
```
* The probability that Salomon improves to three of a kind when the five community cards include two Aces or two Kings
```{r}
Pr_3 <- dhyper(x = 2, N = 48, K = 3, n = 5) * 2
```

Then for each of the first two scenarios, we need to calculate the conditional probability that there is no 10
among the five community cards, in which case Holz would improve to three of a kind (which actually happened).
Thus, the probability that Holz's hand beats Salomon's hand is
```{r}
1 - (Pr_1pair * (1 - dhyper(x = 1, N = 47, K = 2, n = 4)) +
     Pr_2pair * (1 - dhyper(x = 1, N = 46, K = 2, n = 3)) +
     Pr_3)
```
Thus, Holz would be a slight favorite and this conclusion would continue to hold if we were more rigorous about 
accounting for the tiny probability that one or both of their hands ends up being better than three of a kind.
However, this problem ignores the possibility that Kaverman has an Ace, which is an important consideration
that we take up in the next problem.

## Probability that Kaverman has an Ace

This question can be answered just by counting up the proportion of hands that Kaverman would go all-in with
that have an Ace that is not `r card("A", "H")` but do not have `r card("10", "S")` or `r card("10", "C")`. That
leaves

* Big pairs (Js or better), of which there are 21 remaining hands:
    * `r card("A", "S")` `r card("A", "D")`, `r card("A", "S")` `r card("A", "C")`, `r card("A", "D")` `r card("A", "C")`
    * `r card("K", "S")` `r card("K", "H")`, `r card("K", "S")` `r card("K", "D")`, `r card("K", "S")` `r card("K", "C")`,
      `r card("K", "H")` `r card("K", "D")`, `r card("K", "H")` `r card("K", "C")`, `r card("K", "D")` `r card("K", "C")`  
    * `r card("Q", "S")` `r card("Q", "H")`, `r card("Q", "S")` `r card("Q", "D")`, `r card("Q", "S")` `r card("Q", "C")`,
      `r card("Q", "H")` `r card("Q", "D")`, `r card("Q", "H")` `r card("Q", "C")`, `r card("Q", "D")` `r card("Q", "C")`
    * `r card("J", "S")` `r card("J", "H")`, `r card("J", "S")` `r card("J", "D")`, `r card("J", "S")` `r card("J", "C")`,
      `r card("J", "H")` `r card("J", "D")`, `r card("J", "H")` `r card("J", "C")`, `r card("J", "D")` `r card("J", "C")`
* A pair of 10s, of which there is only $1$ combination remaining:
    * `r card("10", "H")` `r card("10", "D")`
* Medium pairs (6s or better), of which there are four pairs $\times$ six suit combinations for 24 hands:
    * `r card("9", "S")` `r card("9", "H")`, `r card("9", "S")` `r card("9", "D")`, `r card("9", "S")` `r card("9", "C")`,
      `r card("9", "H")` `r card("9", "D")`, `r card("9", "H")` `r card("9", "C")`, `r card("9", "D")` `r card("9", "C")`,  
    * `r card("8", "S")` `r card("8", "H")`, `r card("8", "S")` `r card("8", "D")`, `r card("8", "S")` `r card("8", "C")`,
      `r card("8", "H")` `r card("8", "D")`, `r card("8", "H")` `r card("8", "C")`, `r card("8", "D")` `r card("8", "C")`,
    * `r card("7", "S")` `r card("7", "H")`, `r card("7", "S")` `r card("7", "D")`, `r card("7", "S")` `r card("7", "C")`,
      `r card("7", "H")` `r card("7", "D")`, `r card("7", "H")` `r card("7", "C")`, `r card("7", "D")` `r card("7", "C")`,  
    * `r card("6", "S")` `r card("6", "H")`, `r card("6", "S")` `r card("6", "D")`, `r card("6", "S")` `r card("6", "C")`,
      `r card("6", "H")` `r card("6", "D")`, `r card("6", "H")` `r card("6", "C")`, `r card("6", "D")` `r card("6", "C")`
* Ace and at least a nine of the same suit, of which there are 13 remaining hands:
    * `r card("A", "S")` `r card("K", "S")`, `r card("A", "S")` `r card("Q", "S")`, `r card("A", "S")` `r card("J", "S")`,
      `r card("A", "S")` `r card("9", "S")`
    * `r card("A", "D")` `r card("K", "D")`, `r card("A", "D")` `r card("Q", "D")`,
      `r card("A", "D")` `r card("J", "D")`, `r card("A", "D")` `r card("10", "D")`, `r card("A", "D")` `r card("9", "D")`
    * `r card("A", "C")` `r card("K", "C")`, `r card("A", "C")` `r card("Q", "C")`, `r card("A", "C")` `r card("J", "C")`,
      `r card("A", "C")` `r card("9", "C")`
* Ace and a five of the same suit, for any of the three remaining suits: `r card("A", "S")` `r card("5", "S")`,
  `r card("A", "D")` `r card("5", "D")`, `r card("A", "C")` `r card("5", "C")`
* Both cards at least 10 of the same suit, of which there 18 remaining hands:
    * `r card("K", "S")` `r card("Q", "S")`, `r card("K", "S")` `r card("J", "S")`,
      `r card("Q", "S")` `r card("J", "S")`
    * `r card("K", "H")` `r card("Q", "H")`, `r card("K", "H")` `r card("J", "H")`, `r card("K", "H")` `r card("10", "H")`,
      `r card("Q", "H")` `r card("J", "H")`, `r card("Q", "H")` `r card("10", "H")`, `r card("J", "H")` `r card("10", "H")`  
    * `r card("K", "D")` `r card("Q", "D")`, `r card("K", "D")` `r card("J", "D")`, `r card("K", "D")` `r card("10", "D")`,
      `r card("Q", "D")` `r card("J", "D")`, `r card("Q", "D")` `r card("10", "D")`, `r card("J", "D")` `r card("10", "D")` 
    * `r card("K", "C")` `r card("Q", "C")`, `r card("K", "C")` `r card("J", "C")`,
      `r card("Q", "C")` `r card("J", "C")`
* Ace and at least a Jack of different suit, of which there are six intersections $\times {3 \choose 2}$ remaining suits for 27 hands:
    * `r card("A", "S")` `r card("K", "H")`, `r card("A", "S")` `r card("K", "D")`, `r card("A", "S")` `r card("K", "C")`
    * `r card("A", "S")` `r card("Q", "H")`, `r card("A", "S")` `r card("Q", "D")`, `r card("A", "S")` `r card("Q", "C")`
    * `r card("A", "S")` `r card("J", "H")`, `r card("A", "S")` `r card("J", "D")`, `r card("A", "S")` `r card("J", "C")`
    * `r card("A", "D")` `r card("K", "S")`, `r card("A", "D")` `r card("K", "H")`, `r card("A", "D")` `r card("K", "C")`
    * `r card("A", "D")` `r card("Q", "S")`, `r card("A", "D")` `r card("Q", "H")`, `r card("A", "D")` `r card("Q", "C")`  
    * `r card("A", "D")` `r card("J", "S")`, `r card("A", "D")` `r card("J", "H")`, `r card("A", "D")` `r card("J", "C")`
    * `r card("A", "C")` `r card("K", "S")`, `r card("A", "C")` `r card("K", "H")`, `r card("A", "C")` `r card("K", "D")`
    * `r card("A", "C")` `r card("Q", "S")`, `r card("A", "C")` `r card("Q", "H")`, `r card("A", "C")` `r card("Q", "D")`
    * `r card("A", "C")` `r card("J", "S")`, `r card("A", "C")` `r card("J", "H")`, `r card("A", "C")` `r card("J", "D")`
 
Of those $107$ hands, $46$ have at least one Ace, so the conditional probability that Kaverman has an Ace is

$$\frac{46}{107} \approx 0.43$$
Holz could then use this information to compute the probability that Salomon's remaining face-down card is
an Ace given that it is an Ace or a King as

$$\frac{46}{107} \times \frac{2}{6} + \frac{107 - 46}{107} \times \frac{3}{7} \approx 0.388$$
although this does not account for the smaller but non-negligible possibility that Kaverman has a King.

# How Much Expected Money Did Salomon Cost Himself?

The bulk of the expected prize money comes from the roughly $\frac{1}{3}$ chance
Salomon would have of winning first place and its prize of \$10 million
```{r}
stacks <- c(Salomon = 43800000, Holz = 20495000, Bonomo = 51030000, Smith = 19280000)
proportions <- stacks / sum(stacks)
Pr_1 <- proportions["Salomon"]
E <- 10 * Pr_1
```
However, we also have to consider the probability that Salomon comes in second
place, given that someone else comes in first place

$$\Pr\left(\mbox{Salomon second}\right) = 
  \Pr\left(\mbox{Holz, Salomon}\right) + \Pr\left(\mbox{Bonomo, Salomon}\right) +
  \Pr\left(\mbox{Smith, Salomon}\right)$$

or

```{r}
Pr_2 <- proportions["Holz"]   * stacks["Salomon"] / sum(stacks[c("Salomon", "Bonomo", "Smith")]) + 
        proportions["Bonomo"] * stacks["Salomon"] / sum(stacks[c("Salomon", "Holz", "Smith")]) +
        proportions["Smith"]  * stacks["Salomon"] / sum(stacks[c("Salomon", "Holz", "Bonomo")])
```
which is then multiplied by \$6 million
```{r}
E <- E + Pr_2 * 6
```

Similarly, we have to consider the probability that Salomon comes in third
place, given that two other people come in first and second place

$$\Pr\left(\mbox{Salomon third}\right) = 
  \Pr\left(\mbox{Holz, Bonomo, Salomon}\right)  + \Pr\left(\mbox{Bonomo, Holz, Salomon}\right) + \\
  \Pr\left(\mbox{Holz, Smith, Salomon}\right)   + \Pr\left(\mbox{Smith, Holz, Salomon}\right) + \\
  \Pr\left(\mbox{Bonomo, Smith, Salomon}\right) + \Pr\left(\mbox{Smith, Bonomo, Salomon}\right)$$

or in R code  
```{r}
Pr_3 <- proportions["Holz"] * 
        stacks["Bonomo"]  / sum(stacks[c("Salomon", "Bonomo", "Smith")]) * 
        stacks["Salomon"] / sum(stacks[c("Salomon", "Smith")]) + 
  
        proportions["Bonomo"] * 
        stacks["Holz"] / sum(stacks[c("Salomon", "Holz", "Smith")]) * 
        stacks["Salomon"] / sum(stacks[c("Salomon", "Smith")]) +
  
        proportions["Holz"] *
        stacks["Smith"]  / sum(stacks[c("Salomon", "Bonomo", "Smith")]) *
        stacks["Salomon"] / sum(stacks[c("Salomon", "Bonomo")]) + 
  
        proportions["Smith"] * 
        stacks["Holz"] / sum(stacks[c("Salomon", "Bonomo")]) * 
        stacks["Salomon"] / sum(stacks[c("Salomon", "Bonomo")]) +
  
        proportions["Bonomo"] * 
        stacks["Smith"]  / sum(stacks[c("Salomon", "Holz", "Smith")]) *
        stacks["Salomon"] / sum(stacks[c("Salomon", "Holz")]) +
  
        proportions["Smith"] * 
        stacks["Bonomo"]  / sum(stacks[c("Salomon", "Holz", "Bonomo")]) * 
        stacks["Salomon"] / sum(stacks[c("Salomon", "Holz")])
```
which gets multipled by \$4 million
```{r}
E <- E + Pr_3 * 4
```
Finally, we have the complimentary probability that Salomon comes in fourth place
```{r}
Pr_4 <- 1 - proportions["Salomon"] - Pr_2 - Pr_3
```
which gets multiplied by \$2.84 million
```{r}
(E <- E + Pr_4 * 2.84)
```
Altogether, Salomon's expected prize money if he had not exposed the `r card("A", "H")`
would have been \$`r E` million, as opposed to the \$2.84 million he actually won.
