---
title: "Bayesian Probability Lab"
author: "Ben Goodrich"
editor_options:
  chunk_output_type: console
output: 
  html_document: 
    number_sections: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(crayon)
card <- function(value = c("A", as.character(2:9), "10", "J", "Q", "K"), 
                 suit = c("S", "H", "D", "C"),
                 style = NULL) {
  value <- match.arg(value)
  suit <- match.arg(suit)
  
  deck <- list(S = c("\U1F0A1", "\U1F0A2", "\U1F0A3", "\U1F0A4", "\U1F0A5", "\U1F0A6", "\U1F0A7",
                     "\U1F0A8", "\U1F0A9", "\U1F0AA", "\U1F0AB", "\U1F0AD", "\U1F0AE"),
               H = c("\U1F0B1", "\U1F0B2", "\U1F0B3", "\U1F0B4", "\U1F0B5", "\U1F0B6", "\U1F0B7",
                     "\U1F0B8", "\U1F0B9", "\U1F0BA", "\U1F0BB", "\U1F0BD", "\U1F0BE"),
               D = c("\U1F0C1", "\U1F0C2", "\U1F0C3", "\U1F0C4", "\U1F0C5", "\U1F0C6", "\U1F0C7",
                     "\U1F0C8", "\U1F0C9", "\U1F0CA", "\U1F0CB", "\U1F0CD", "\U1F0CE"),
               C = c("\U1F0D1", "\U1F0D2", "\U1F0D3", "\U1F0D4", "\U1F0D5", "\U1F0D6", "\U1F0D7",
                     "\U1F0D8", "\U1F0D9", "\U1F0DA", "\U1F0DB", "\U1F0DD", "\U1F0DE"))
               
  value <- switch(value,
                  "A" = 1L,
                  "10" = 10L,
                  "J" = 11L,
                  "Q" = 12L,
                  "K" = 13L,
                  as.integer(value))
  if (is.null(style)) style <- if (suit %in% c("H", "D")) red else reset
  return(fansi::sgr_to_html(style(deck[[suit]][value])))
}
```

In this lab, we will be considering a hand from a poker tournament that you should watch at

https://youtu.be/dZyGfk2HDaY

The goals of this lab are to

1. See how professional poker players update their beliefs about their opponents' face-down
  cards as they accumulate more data
2. Understand how decisions can be made to maximize _expected_ utility
3. Calculate probabilities, in this case using the R language

Before starting, you need to know just a little bit about the applicable rules of poker,
specifically the most popular form of poker No Limit Texas Hold'em. Each player at the
table is dealt two face-down cards from a shuffled deck that consists of $52$ cards whose $13$
values (in increasing order) are 2 3 4 5 6 7 8 9 10 J Q K A on each of four suits
(spades, hearts, diamonds, and clubs). Thus, there are ${52 \choose 2} = \frac{52!}{2! \times 50!} 
= 1326$ possible combinations of two face-down cards, as their order is not important. The players 
can look at their own two face-down cards and must combine them with three of the five face-up cards 
in the middle of the table to form a five-card poker hand. More improbable five-card poker hands beat 
less improbable hands, and if two or more players have equally improbable hands, the player with the 
highest card wins. Specifically,

* A "straight flush" consists of five cards that have consecutive values and are all of
  the same suit, which is rare but beats all other hands
* "Four of a kind", consists of four cards that all have the same value and one other
  irrelevant card, which beats all hands but a "straight flush"
* A "full house", consists of three cards that all have the same value and two other
  cards that also have the same value, which beats all hands except "four of a kind"
  and a "straight flush"
* A "flush" consists of five cards that all have the same suit, which beats all
  hands not previously mentioned
* A "straight" consists of five cards that have consecutive values. Note that an Ace
  can be considered higher than a King or lower than a 2 for the purpose of forming
  a straight, so A K Q J 10 is a straight, as is A 2 3 4 5. A "straight"
  beats all hands not previously mentioned.
* "Three of a kind", consists of three cards that all have the same value and two
  other irrelevant cards, which beats all hands not previously mentioned
* "Two pair" consists of two cards that have the same value, two other cards that
  also have the same value, and a fifth irrelevant card. Again, having "two pair"
  beats all hands not previously mentioned.
* "One pair" consists of two cards that have the same value and three irrelevant
  cards, which beats any hand that does not have (at least) two cards of the same 
  value or a straight or a flush.
  
If none of the players has any of the aforementioned hands, the player with the
highest card among the two face-down cards wins all the poker chips that have
been bet.

There are up to four rounds of betting, one before any cards in the middle have
been revealed and three more rounds as cards in the middle are incrementally
revealed. The players take turns clockwise around the table and each has the 
opportunity to 

* "Call", which means to match the number of poker chips bet by the previous player
* "Raise", which means to bet in excess of the number of poker chips bet by the
  previous player
* "Fold", which means to bet no additional poker chips and forgo the opportunity to
  win any of the chips that have been bet this hand. At that point, the player hands
  their two face-down cards back to the dealer.
  
In addition, before any cards are dealt, each of the five players has to bet some (in this case
$120,000$) poker chips as an "ante", one player has to bet a lot (in this case
$600,000$) poker chips as the "big blind", and one player has to bet half as many (in this case
$300,000$) poker chips as the "small blind". Thus, in this case, there are
$1,500,000$ poker chips that have been bet before any of the players see their
two face-down cards. For more details on the rules, you can refer to (although it is not really
necessary)

https://en.wikipedia.org/wiki/Texas_hold_%27em

At the outset of this video, the five players have the following amounts of poker chips:
```{r}
stacks <- c(Salomon  = 27450000, # must post the "big blind" of 600,000 and ante 120,000
            Kaverman =  8025000, # has to start the betting and ante 120,000
            Holz     = 28675000, # must ante 120,000
            Bonomo   = 51150000, # must ante 120,000
            Smith    = 19700000) # must post the "small blind" of 300,000 and ante 120,000
```
However, in a poker tournament, you cannot simply leave the table and redeem your poker
chips for cash. Rather, poker players are eliminated from the tournament when they have
no poker chips left and cash prizes are awarded for being among the last few players to
be eliminated. Specifically, in this video, the cash prizes were

* 5th place: $2 million
* 4th place: $2.84 million
* 3rd place: $4 million
* 2nd place: $6 million
* 1st place: $10 million

and nothing was awarded for 6th place or worse. Just prior to this video, the 6th place
player was eliminated.

# Byron Kaverman's Decision

## Prior

Byron Kaverman is dealt `r card("A", "C")` `r card("5", "C")`. **What was the _ex ante_
probability of being dealt these two cards (or really any combination of two cards)?**

## Posterior

Kaverman starts the video by betting all of his remaining poker chips, which is
standard for a poker player in this situation, namely one who has a good hand but much fewer
poker chips than all the other players at the table. Specifically, simulations have shown
that for someone in Kaverman's situation, it is better (albeit high variance) to bet all of one's poker 
chips than to fold (or to bet only some poker chips) if and only if they have any of the following 
$138$ hands:

* Big pairs (Js or better), of which there are four pairs $\times$ six suit combinations for 24 hands:
    * `r card("A", "S")` `r card("A", "H")`, `r card("A", "S")` `r card("A", "D")`, `r card("A", "S")` `r card("A", "C")`,
      `r card("A", "H")` `r card("A", "D")`, `r card("A", "H")` `r card("A", "C")`, `r card("A", "D")` `r card("A", "C")`
    * `r card("K", "S")` `r card("K", "H")`, `r card("K", "S")` `r card("K", "D")`, `r card("K", "S")` `r card("K", "C")`,
      `r card("K", "H")` `r card("K", "D")`, `r card("K", "H")` `r card("K", "C")`, `r card("K", "D")` `r card("K", "C")`  
    * `r card("Q", "S")` `r card("Q", "H")`, `r card("Q", "S")` `r card("Q", "D")`, `r card("Q", "S")` `r card("Q", "C")`,
      `r card("Q", "H")` `r card("Q", "D")`, `r card("Q", "H")` `r card("Q", "C")`, `r card("Q", "D")` `r card("Q", "C")`
    * `r card("J", "S")` `r card("J", "H")`, `r card("J", "S")` `r card("J", "D")`, `r card("J", "S")` `r card("J", "C")`,
      `r card("J", "H")` `r card("J", "D")`, `r card("J", "H")` `r card("J", "C")`, `r card("J", "D")` `r card("J", "C")`
* A pair of 10s, of which there are 6 suit combinations:
    * `r card("10", "S")` `r card("10", "H")`, `r card("10", "S")` `r card("10", "D")`, `r card("10", "S")` `r card("10", "C")`,
      `r card("10", "H")` `r card("10", "D")`, `r card("10", "H")` `r card("10", "C")`, `r card("10", "D")` `r card("10", "C")`
* Medium pairs (6s or better), of which there are four pairs $\times$ six suit combinations for 24 hands:
    * `r card("9", "S")` `r card("9", "H")`, `r card("9", "S")` `r card("9", "D")`, `r card("9", "S")` `r card("9", "C")`,
      `r card("9", "H")` `r card("9", "D")`, `r card("9", "H")` `r card("9", "C")`, `r card("9", "D")` `r card("9", "C")`,  
    * `r card("8", "S")` `r card("8", "H")`, `r card("8", "S")` `r card("8", "D")`, `r card("8", "S")` `r card("8", "C")`,
      `r card("8", "H")` `r card("8", "D")`, `r card("8", "H")` `r card("8", "C")`, `r card("8", "D")` `r card("8", "C")`,
    * `r card("7", "S")` `r card("7", "H")`, `r card("7", "S")` `r card("7", "D")`, `r card("7", "S")` `r card("7", "C")`,
      `r card("7", "H")` `r card("7", "D")`, `r card("7", "H")` `r card("7", "C")`, `r card("7", "D")` `r card("7", "C")`,  
    * `r card("6", "S")` `r card("6", "H")`, `r card("6", "S")` `r card("6", "D")`, `r card("6", "S")` `r card("6", "C")`,
      `r card("6", "H")` `r card("6", "D")`, `r card("6", "H")` `r card("6", "C")`, `r card("6", "D")` `r card("6", "C")`
* Ace and at least a nine of the same suit, of which there are five intersections $\times$ four suits for 20 hands:
    * `r card("A", "S")` `r card("K", "S")`, `r card("A", "S")` `r card("Q", "S")`, `r card("A", "S")` `r card("J", "S")`,
      `r card("A", "S")` `r card("10", "S")`, `r card("A", "S")` `r card("9", "S")`
    * `r card("A", "H")` `r card("K", "H")`, `r card("A", "H")` `r card("Q", "H")`, `r card("A", "H")` `r card("J", "H")`,
      `r card("A", "H")` `r card("10", "H")`, `r card("A", "H")` `r card("9", "H")`
    * `r card("A", "D")` `r card("K", "D")`, `r card("A", "D")` `r card("Q", "D")`,
      `r card("A", "D")` `r card("J", "D")`, `r card("A", "D")` `r card("10", "D")`, `r card("A", "D")` `r card("9", "D")`
    * `r card("A", "C")` `r card("K", "C")`, `r card("A", "C")` `r card("Q", "C")`, `r card("A", "C")` `r card("J", "C")`,
      `r card("A", "C")` `r card("10", "C")`, `r card("A", "C")` `r card("9", "C")`
* Ace and a five of the same suit, for any of the four suits: `r card("A", "S")` `r card("5", "S")`,
  `r card("A", "H")` `r card("5", "H")`, `r card("A", "D")` `r card("5", "D")`, `r card("A", "C")` `r card("5", "C")`
* Both cards at least 10 of the same suit, of which there are six intersections $\times$ four suits for 24 hands:
    * `r card("K", "S")` `r card("Q", "S")`, `r card("K", "S")` `r card("J", "S")`, `r card("K", "S")` `r card("10", "S")`,
      `r card("Q", "S")` `r card("J", "S")`, `r card("Q", "S")` `r card("10", "S")`, `r card("J", "S")` `r card("10", "S")`
    * `r card("K", "H")` `r card("Q", "H")`, `r card("K", "H")` `r card("J", "H")`, `r card("K", "H")` `r card("10", "H")`,
      `r card("Q", "H")` `r card("J", "H")`, `r card("Q", "H")` `r card("10", "H")`, `r card("J", "H")` `r card("10", "H")`  
    * `r card("K", "D")` `r card("Q", "D")`, `r card("K", "D")` `r card("J", "D")`, `r card("K", "D")` `r card("10", "D")`,
      `r card("Q", "D")` `r card("J", "D")`, `r card("Q", "D")` `r card("10", "D")`, `r card("J", "D")` `r card("10", "D")` 
    * `r card("K", "C")` `r card("Q", "C")`, `r card("K", "C")` `r card("J", "C")`, `r card("K", "C")` `r card("10", "C")`,
      `r card("Q", "C")` `r card("J", "C")`, `r card("Q", "C")` `r card("10", "C")`, `r card("J", "C")` `r card("10", "C")`
* Ace and at least a Jack of different suit, of which there are six intersections $\times {4 \choose 2}$ suits for 36 hands:
    * `r card("A", "S")` `r card("K", "H")`, `r card("A", "S")` `r card("K", "D")`, `r card("A", "S")` `r card("K", "C")`
    * `r card("A", "S")` `r card("Q", "H")`, `r card("A", "S")` `r card("Q", "D")`, `r card("A", "S")` `r card("Q", "C")`
    * `r card("A", "S")` `r card("J", "H")`, `r card("A", "S")` `r card("J", "D")`, `r card("A", "S")` `r card("J", "C")`
    * `r card("A", "H")` `r card("K", "S")`, `r card("A", "H")` `r card("K", "D")`, `r card("A", "H")` `r card("K", "C")`  
    * `r card("A", "H")` `r card("Q", "S")`, `r card("A", "H")` `r card("Q", "D")`, `r card("A", "H")` `r card("Q", "C")`
    * `r card("A", "H")` `r card("J", "S")`, `r card("A", "H")` `r card("J", "D")`, `r card("A", "H")` `r card("J", "C")`
    * `r card("A", "D")` `r card("K", "S")`, `r card("A", "D")` `r card("K", "H")`, `r card("A", "D")` `r card("K", "C")`
    * `r card("A", "D")` `r card("Q", "S")`, `r card("A", "D")` `r card("Q", "H")`, `r card("A", "D")` `r card("Q", "C")`  
    * `r card("A", "D")` `r card("J", "S")`, `r card("A", "D")` `r card("J", "H")`, `r card("A", "D")` `r card("J", "C")`
    * `r card("A", "C")` `r card("K", "S")`, `r card("A", "C")` `r card("K", "H")`, `r card("A", "C")` `r card("K", "D")`
    * `r card("A", "C")` `r card("Q", "S")`, `r card("A", "C")` `r card("Q", "H")`, `r card("A", "C")` `r card("Q", "D")`
    * `r card("A", "C")` `r card("J", "S")`, `r card("A", "C")` `r card("J", "H")`, `r card("A", "C")` `r card("J", "D")`

Professional poker players do not look at their two face-down cards until it is their turn to bet in order to not
give away any information with their body language. **At the instant that Kaverman goes all-in but before anyone else
looks at their two face-down cards, from another player's perspective, what is the posterior probability that Kaverman
has `r card("A", "C")` `r card("5", "C")`? Can you write this posterior probability as a special case of Bayes' Rule?**

# Fedor Holz's First Decision

Fedor Holz is dealt `r card("10", "S")` `r card("10", "C")` and is next to bet.

## Probability of Three of a Kind

**If Kaverman had a pair of Jacks, Queens, Kings, or Aces, what is the probability that Holz's five-card hand ends up
being better than Kaverman's five-card hand?** For the purposes of this question, you can ignore the tiny probability
that either player ends up with a straight flush, four of a kind, full house, flush, or straight.

## Probability of Trailing

Holz's hand would be a favorite to beat Kaverman's hand unless Kaverman has a pair of Jacks, Queens, Kings, or Aces
(and they would almost certainly be tied if Kaverman had `r card("10", "H")` `r card("10", "D")`). **From Holz's
perspective at the time he calls Kaverman, what is the probability Holz's hand is not a favorite to beat Kaverman's
hand?**

# Rick Salomon's Decision

After Holz calls Kaverman's bet and both Justin Bonomo and Dan Smith fold, $17,550,000$ chips have been bet. Rick
Salomon is dealt `r card("A", "H")` `r card("K", "H")` and bets all of his remaining $26,730,000$ chips. One of
the commentators says about 1:38 into the video that Salomon would not do so unless he had a pair of Jacks,
Queens, Kings, Aces, or some combination of an Ace and a King (that may or may not be of the same suit). **Use
probability to inexactly sketch an argument as to why that is true.**

# Fedor Holz's Second Decision

Between 1:03 and 1:04 of the video, Salomon picks up the `r card("A", "H")` in his right hand in such a way
that everyone (including the referee, who is wearing a tuxedo) can see it. By rule, if a player's face-down
card is revealed to any other player, it has to be turned face-up (in order to prevent players from colluding).
At this point, Holz knows that Salomon's remaining face-down card is one of `r card("A", "S")`, `r card("A", "D")`,
`r card("A", "C")`, `r card("K", "S")`, `r card("K", "H")`, `r card("K", "D")`, or `r card("K", "C")`.

## Beating a King that is not `r card("K", "H")`

Suppose Salomon's remaining face-down card is a King but not `r card("K", "H")`. **What is the probability that 
Holz's five-card hand ends up being better than Salomon's five-card hand?** For the purposes of this question, 
you can ignore the tiny probability that either player ends up with a straight flush, four of a kind, full house, 
flush, or straight. Also, you can ignore the (considerable) possibility that Kaverman has an Ace and / or a King.

## Probability that Kaverman has an Ace

Starting at about 2:52 into the video, one of the commentators rambles that the proposition that Kaverman
has an Ace is "one of those 50-50 things" and later "maybe there is a 30% chance that [Kaverman] has an
Ace but that influence's [Holz's] decision". **From Holz's perspective, what is the probability that
Kaverman has at least one Ace?**

# How Much Expected Money Did Salomon Cost Himself?

Suppose Salomon did not expose  `r card("A", "H")` as he went all-in, in which case Holz would have folded 
and Kaverman would have been eliminated when Salomon made two pair. The four remaining amounts of poker chips 
would then be
```{r}
stacks <- c(Salomon = 43800000, Holz = 20495000, Bonomo = 51030000, Smith = 19280000)
```
Under the Independent Chip Model (ICM), the probability of each remaining player winning a poker 
tournament is equal to the proportion of chips that they have:
```{r}
(proportions <- stacks / sum(stacks))
```
Then, the probability of a player coming in second place, given that they do not come in first
place, is equal to the proportion of chips the player has *excluding the chips of the player who wins
first place*. When calculating the probability of a player coming in second place, you have to
take into account that there are three different bivariate probabilities that the player comes
in second place and another player comes in first place. The probability of a player coming in 
third place, given that they do not come in first place or second place, is conceptually 
similar. **Under the above assumptions, what would be the expected amount of prize money for
Salomon, as compared to the $2.84 million he actually won for fourth place?**
